The Calculus of Linear Constructions

### 2021-12-02 writing/draft5/draft5.tex
writing

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mdwtab}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{float}
\usepackage{tikz-cd}
\usepackage{enumitem}

@@

\subsection{Linear Connectives}
In Figure \ref{connectives}, we demonstrate how the $\otimes$ and $\oplus$ connectives of linear logic could be defined in CILC as inductive linear types. For any term with inductive linear type, linearity will ultimately force the term to be eliminated by a Case expression.

\begin{figure}[h]
  \caption{Inductively Defined Linear Connectives}
  \begin{align*}
    \_\otimes\_ :=\
     & \ind{L}(X : (x \utype L_i) \rightarrow (y \utype L_i) \rightarrow L)\{        \\
     & \sep\ (A \utype L_i) \rightarrow (B \utype L_i)                               \\
     & \qquad \rightarrow (\_ \ltype A)  \multimap (\_ \ltype B) \multimap (X\ A\ B) \\
     & \}
    \\
    \lrangle{\_,\_} := \
     & \constr(1, \otimes)
    \\\\
    \_\oplus\_ := \
     & \ind{L}(X : (x \utype L_i) \rightarrow (y \utype L_i) \rightarrow L)\{        \\
     & \sep \ (A \utype L_i) \rightarrow (B \utype L_i)
    \rightarrow (\_ \ltype A)  \multimap (X\ A\ B)                                   \\
     & \sep \ (A \utype L_i) \rightarrow (B \utype L_i)
    \rightarrow (\_ \ltype B)  \multimap (X\ A\ B)                                   \\
     & \}
    \\
    \inl(\_) := \
     & \constr(1, \oplus)
    \\
    \inr(\_) := \
     & \constr(2, \oplus)
  \end{align*}
  \Description{}
  \label{connectives}
\end{figure}

>>>>

% packages
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mdwtab}
\usepackage{syntax}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{float}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{minted}

@@

\subsection{Non-Linear Connectives}
Non-linear inductive types have been extensively studied since their inception. We will omit discussion of non-linear inductive types here as all inductive definitions in the predicative fragment of CIC are available in CILC.

\subsection{Linear Connectives}
In Figure \ref{lconnectives}, we demonstrate how the $\otimes$ and $\oplus$ connectives of linear logic could be defined in CILC as inductive linear types. For any term with inductive linear type, linearity will ultimately force the term to be eliminated by a Case expression.

\begin{figure}[h]
  \caption{Inductively Defined Linear Connectives}
  \begin{align*}
    \_\otimes\_ :=\
     & \ind{L}(X : (A \utype L_i) \rightarrow (B \utype L_i) \rightarrow L_i)\{      \\
     & \sep\ (A \utype L_i) \rightarrow (B \utype L_i)                               \\
     & \qquad \rightarrow (\_ \ltype A)  \multimap (\_ \ltype B) \multimap (X\ A\ B) \\
     & \}
    \\
    \lrangle{\_,\_} := \
     & \constr(1, \otimes)
    \\\\
    \_\oplus\_ := \
     & \ind{L}(X : (A \utype L_i) \rightarrow (B \utype L_i) \rightarrow L_i)\{      \\
     & \sep \ (A \utype L_i) \rightarrow (B \utype L_i)
    \rightarrow (\_ \ltype A)  \multimap (X\ A\ B)                                   \\
     & \sep \ (A \utype L_i) \rightarrow (B \utype L_i)
    \rightarrow (\_ \ltype B)  \multimap (X\ A\ B)                                   \\
     & \}
    \\
    \inl(\_) := \
     & \constr(1, \oplus)
    \\
    \inr(\_) := \
     & \constr(2, \oplus)
  \end{align*}
  \Description{}
  \label{lconnectives}
\end{figure}

\subsection{Mixed Connectives}
In Figure \ref{mconnectives}, we demonstrate how the $F$ connective of LNL$_D$ \cite{neel15} can be expressed as an inductive linear type. The interesting property of $F$ is that the type of its linear component is dependent on the non-linear component.

\begin{figure}[h]
  \caption{Linear Non-Linear Connective}
  \begin{align*}
    F := \
     & \ind{L}(X : (A \utype U_i) \rightarrow
    (B \utype (\_ \utype A) \rightarrow L_i) \rightarrow L_i)\{ \\
     & \sep \ (A \utype U_i) \rightarrow
    (B \utype (\_ \utype A) \rightarrow L_i)                    \\
     & \qquad \rightarrow (m \utype A) \rightarrow
    (\_ \ltype B\ m) \multimap (X\ A\ B)                        \\
     & \}
    \\
    [\_,\_] :=\
     & \constr(1, F)
  \end{align*}
  \Description{}
  \label{mconnectives}
\end{figure}

\section{Examples}
We shall write examples in the concrete syntax of our implementation. We deliberately chose to model our concrete syntax after Coq such that it would be easier for readers already familiar with Coq to adapt to our syntax.

\subsection{Stateful Reasoning}
We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator.
\begin{definition}
  State axioms.
  \begin{itemize}
    \item $l \mapsto_A m : \mathbb{N} \rightarrow (A : U) \rightarrow A \rightarrow L$ \\
          The $\mapsto$ is a linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, it is a proof that a term $m$ of non-linear type $A$ is currently stored at address $l$.
    \item $\textbf{new} : (A : U) \rightarrow (m : A) \rightarrow F l:\mathbb{N}.(l \mapsto_A m)$ \\
          $\new$ is a function that allocates a new memory cell for term $m$ of non-linear type $A$. It results in an address $l$ paired with an assertion $(l \mapsto_A m)$.
    \item $\textbf{free} : (A : U) \rightarrow F l:\mathbb{N}.F m:A.(l \mapsto_A m) \rightarrow \unit$ \\
          $\free$ is a function that de-allocates the memory cell at location $l$ which is currently in use. This can be seen in its input $F l:\mathbb{N}.F m:A.(l \mapsto_A m)$, where the assertion component $(l \mapsto_A m)$ is a proof that location $l$ is currently allocated.
    \item $\textbf{get} : (A : U) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \rightarrow F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$ \\
          $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto_A m)$ that there is indeed a term $m$ currently stored there. The type of $\get$'s output is the rather complicated $F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$. This not only returns the term stored at $l$, but also a proof that the returned term is exactly the same as what $(l \mapsto_A m)$ claimed to have stored. The proof $(l \mapsto_A m)$ is returned unchanged as well.
    \item $\textbf{set} : (A\ B : U) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \multimap (n : B) \multimap (l \mapsto_B n)$ \\
          $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term $m$ of type $A$, $\set$ may put term $n$ of type $B$ into the the store at address $l$, overwriting the original term. This is the so called "strong update" operation. The consumption of $(l \mapsto_A m)$ and return of $(l \mapsto_B n)$ accurately characterize this update process.
  \end{itemize}
\end{definition}
Due to the fact that $(l \mapsto_A m)$ is a linear type, proofs of this type can not be duplicated and may only be consumed by other state axioms. This is the same non-sharing principle that empowers Separation Logic. Unlike Separation Logic, $(l \mapsto_A m)$ itself is a first-class object in CILC, meaning that it can be the result of computation (large eliminations), input to computations, stored in data structures, etc.

\subsection{Internalization of Effect Laws}
The combination of linear types and dependent types within the state axioms enable the internalization of the so called "effect laws" for state, where these laws are now provable theorems within the CLC framework.

\begin{enumerate}
  \item Performing $\get$ twice on the same address will produce two terms that are provably equal.
  \item After using $\set$ to update the store at an address $l$ to term $m$, performing $\get$ again on the same address will retrieve $m$.
\end{enumerate}

Figure \ref{law1} is an excerpt taken from our implementation. It demonstrates the internalization of Law 1. The usage of implicit parameters allows one to omit writing much of the redundant arguments that could be inferred from context. Here, $\new$ allocates a store for natural number 1, creating a proof $c$ of the form $(l \mapsto_\mathbb{N} 1)$. When $\get$ is applied to $c$ twice, the equality proofs \textit{pf1} $: (1 =_\mathbb{N} x)$ and \textit{pf2} $: (1 =_\mathbb{N} y)$ generated by $\get$ allow one to create proof \textit{pf} $: (x =_\mathbb{N} y)$ that the retrieved values $x$ and $y$ are equal. Finally, due to $c$ having linear type, it is required for it to be freed at the very end.

\begin{figure}[h]
  \caption{Law 1}
  \begin{minted}{coq}
  let [ _, c ] := new _ 1 in
  let [ xeq, c ] := get _ _ _ c in
  let [ yeq, c ] := get _ _ _ c in
  let ( x, pf1 ) := xeq in
  let ( y, pf2 ) := yeq in
  let pf1 := Eq_sym _ _ _ pf1 in
  let pf : Eq _ x y := 
    Eq_trans _ _ _ _ pf1 pf2 
  in free _ _ _ c.
  \end{minted}
  \label{law1}
  \Description{}
\end{figure}

Figure \ref{law2} demonstrates the internalization of Law 2. Again, implicit parameters fill in the annoying bureaucratic arguments whenever they can be inferred. Here, a store is allocated by $\new$ for natural number 1, creating a proof $c$ of the form $(l \mapsto_\mathbb{N} 1)$. After using $\set$ to update the store at $l$ to 2 and then performing $\get$ on $l$, one can create proof \textit{pf} $: (2 =_\mathbb{N} z)$ that the value $z$ retrieved by $\get$ is exactly equal to 2.

\begin{figure}[h]
  \caption{Law 2}
  \begin{minted}{coq}
  let [ _, c ] := new _ 1 in
  let c := set _ _ _ _ c 2 in
  let [ zeq, c ] := get _ _ _ c in
  let ( z, pf1 ) := zeq in
  let pf : Eq _ 2 z := pf1 in
  free _ _ _ c.
  \end{minted}
  \label{law2}
  \Description{}
\end{figure}

\subsection{Protocol Enforcement}
The inclusion of dependent types and linear types in CILC makes it extremely expressive when encoding communication protocols. By indexing each channel with a protocol, CLC can enforce communication on a channel to strictly adhere to its specified protocol.

In Figure \ref{session}, we declare an inductive \texttt{session} type for specifying protocols. Intuitively, the \texttt{session}  forms a stack of possible operations executed in a protocol. The $\SEND$ constructor takes in a non-linear type $A$ and a session $ss$ as its argument, indicating that after sending a message of type $A$, the protocol progresses to stage $ss$. Likewise, the $\RECV$ constructor takes in a non-linear type $A$ and a session $ss$ as its argument, indicating that after receiving a message of type $A$, the protocol progresses to stage $ss$. Finally, the $\END$ constructor indicates that the protocol has finished and communication is over.

\begin{figure}[h]
  \caption{Inductive Session Type}
  \begin{minted}{coq}
  Inductive session : U :=
  | SEND : U -> session -> session
  | RECV : U -> session -> session
  | END : session.
  \end{minted}
  \label{session}
  \Description{}
\end{figure}

In Definition \ref{comm}, we postulate five axioms for communication using the $\session$ type. These axioms could be viewed as an interface to trusted communication libraries.

\begin{definition}
  Communication axioms.
  \begin{itemize}
    \item $\textbf{channel} : \session \rightarrow L$ \\
          The $\channel$ axiom is a type constructor. It is used to form the types of communication channels that obey the protocol specified by its session.
    \item $\textbf{open} : (ss : \session) \rightarrow \channel\ ss$ \\
          When given a protocol, $\open$ creates a channel indexed by this protocol.
    \item $\textbf{close} : \channel\ \END \rightarrow \unit$ \\
          Once all communications specified on a given channel has finished, $\close$ will close and reclaim the channel.
    \item $\textbf{send} : (A : U) \rightarrow A \rightarrow (ss : \session) \rightarrow$ \\
          \phantom{send : (A :)}
          $\channel\ (\SEND\ A\ ss) \rightarrow \channel\ ss$ \\
          After a channel has reached the point in its protocol where a message of type $A$ should be sent, $\send$ may send a term of type $A$ on this channel, causing the channel to progress to the next stage of its protocol.
    \item $\textbf{recv} : (A : U) \rightarrow (ss : \session) \rightarrow $ \\
          \phantom{recv : (A :)}
          $\channel\ (\RECV\ A\ ss) \rightarrow F \_:A.\channel\ ss$
          After a channel has reached the point in its protocol where a message of type $A$ should be received, $\recv$ may receive such a term of type $A$ from the channel, causing the channel to progress to the next stage of its protocol.
  \end{itemize}
  \label{comm}
\end{definition}

The important detail to notice here is that each channel created by $\open$ is of linear type $(\channel\ ss)$ where $ss$ is some arbitrary protocol, hence must be used exactly once. However, the $\send$ and $\recv$ operations return back channels after consuming them, progressing their protocols one stage forward. Coupled with the fact that the $\close$ operation is only able to close channels whose protocols have ended, this forces channels to communicate strictly by their specified protocols.

Dependent types also allows for more precise specifications of protocols themselves. A simple example would be the specification of message sizes. Figure \ref{communication} shows such an example. The inductive type \texttt{ilist} is a length indexed list. Using \texttt{ilist}, we construct a protocol \texttt{ss} that first sends two natural numbers, then receives eight natural numbers and finally sends back a boolean acknowledgement.

\begin{figure}[h]
  \caption{Sized Communication}
  \begin{minted}{coq}
  Inductive ilist (A : U) : Nat -> U :=
  | nil  : ilist A 0
  | cons : A -> (n : Nat) -> 
             ilist A n -> ilist A (S n).

  let ss := 
    SEND (ilist Nat 2) 
      (RECV (ilist Nat 8) (SEND Bool END)) 
  in
  let send_msg : ilist Nat 2 := 
    (cons 1 _ (cons 2 _ nil)) 
  in
  let ch := open ss in
  let ch := send _ send_msg _ ch in
  let [ recv_msg, ch ] := recv _ _ ch in
  let ch := send _ true _ ch in
  close ch.            
  \end{minted}
  \label{communication}
  \Description{}
\end{figure}

\section{Related Work}
Linear types are a class of type systems inspired by Girard' substructural Linear Logic \cite{girard}. Girard notices that the weakening and contraction rules of Classical Logic when restricted carefully, give rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} then applies an analogous restriction to variable usage in simple type theory, leading to the development of linear type theory where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications.

Over the years, work has been done to enrich linear type theories with dependent types. Cervesato and Pfenning extend the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} presents a linear dependent type theory, with syntax and semantics drawing inspiration from DILL \cite{dill}.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's earlier work on mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming in the style of Hoare Type Theory \cite{htt}. Luo et al. \cite{luo} introduce the property of essential linearity and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

\section{Future Work}
The integration of linear types and dependent types opens the door to many topics for research, both in theory and in its applications.

On the theoretical side, we intend to investigate the semantic models of CLC. In early versions of CLC, an impredicative $Prop$ universe sat at the bottom of the universe hierarchy. This has since been removed due to the additional difficulty of working with $Prop$ in the soundness proof of CILC. We believe that these difficulties are not intrinsic and $Prop$ could be added back with careful proof work.

On the application side, we intend to bolster our implementation with more features such as compilation to executable binary, making it a full-featured theorem prover/programming language. We aim to fully verify the typechecking algorithm employed by the current implementation. The unification algorithm used for implicit argument resolution is extremely ad-hoc, as the problem of unification for linear types has been not thoroughly studied, we hope to improve this situation as well.
